---
title:
  "LifesavR Cheatsheet (Introduction to R course) "
author: Ben Whalley
output:
  webex::html_clean
---


<meta property="og:description" content="Included on this page are short examples of everything we teach students in stage 1 to 4, including the project option workshops, plus some extensions/additions which you may find useful to introduce project students to. Extensions (i.e. parts not taught in the programme) are clearly marked." />


> This page is intended as a quick reference when helping students with their data handling and analysis.
>
> Included on this page are short examples of everything we teach students in the LifeSavR introductory/revision course.



# Basic R statements

```{r}
# assign the current year to a variable
current_year <- 2021

# print the variable we created
# (it should display below the chunk, or in the console pane)
current_year
```


```{r}
# do a calculation to create a new variable from an existing one
next_year <- current_year + 1

next_year
```


```{r}
# multiply two numbers
2 * 221
```



# Loading packages

These are the packages used in this cheatsheet (and the lifesavR course:

```{r}
library(tidyverse)
library(psydata)
library(BayesFactor)
library(corrr)
```



# Looking at data

Show only the first 6 rows of the `fuel` data

```{r}
head(fuel)
```


Shows a list of columns in the `development` dataset plus the first few datapoints (as many as will fit)

```{r}
glimpse(development)
```



# Filtering datasets

The `filter()` function selects rows from a dataset:

```{r, eval=F}
# filter on a categorical variable
# == means 'match exactly'
development %>% 
  filter(country == "Kenya")

# this doesn't match any rows because the data spells Kenya with a capital 'K'
development %>% 
  filter(country == "kenya")

# filtering on a numeric variable
development %>% filter(year > 2000)


# combining multiple filters makes things more specific
development %>%
  filter(country == "Kenya") %>%
  filter(year > 2000)

# assign filtered data to a new variable for re-use later on
# if you don't store the data, it is displayed but then lost
development.europe <- development %>% 
  filter(continent == "Europe") %>% 
  filter(year > 1980)

# show the stored subset
development.europe

# make a plot with the subset
development.europe %>% 
    ggplot(aes(gdp_per_capita, life_expectancy)) + 
  geom_point()
```




# Frequencies

```{r}
# this gives us the total number of rows
earnings %>% count()

# numbers of men and women
earnings %>%
  count(gender)

# We can do arithmetic on these numbers
# calculate percentage
2353 / 4483 * 100


# counts for multiple subgroups
earnings %>%
  count(job, gender)
```

```{r}
# making a frwquency plot
earnings %>%
  ggplot(aes(gender)) +
  stat_count()
```



# Summarise


```{r, eval=FALSE}
# use function mean() to summarise mpg column
# the result is a data.frame with a single column named 'mean_mpg'
fuel %>%
  summarise(mean_mpg = mean(mpg))

# if you omit the column name, the summarised column is named after the summary function
# this produces column names which can be awkward to process later in a pipeline
# for example:
fuel %>%
  summarise(mean(mpg))

# median mpg (rather than mean)
fuel %>%
  summarise(median_mpg = median(mpg))

# standard deviation
fuel %>%
  summarise(sd_mpg = sd(mpg))

# summarise two columns at once 
# the mean and sd functions are separated with a comma
# the resulting data frame is stored in a variable called fuel_summary
fuel_summary <- fuel %>%
  summarise(m = mean(mpg), sd = sd(mpg))

# see the stored summary
fuel_summary

# remember, this summary is still a new dataset; you could do more processing on it in extra steps if needed

# combining filter() and summarise()
development %>% 
  filter(continent=="Asia") %>% 
  summarise(Average_life_expectancy = mean(life_expectancy))
```








# Group by

```{r, eval=F}
# calculate mean weight loss in each group in a laborious way
funimagery %>%
  filter(intervention=="MI") %>%
  summarise(mean(weight_lost_end_trt))

funimagery %>%
  filter(intervention=="FIT") %>%
  summarise(mean(weight_lost_end_trt))


# use group_by to split the data and summarise each group
funimagery %>%
  group_by(intervention) %>%
  summarise(mean(weight_lost_end_trt))

```


```{r}
# example of grouping by two columns at once
funimagery %>%
  group_by(gender, intervention) %>%
  summarise(mean(weight_lost_end_trt))

```


Calculate the mean and SD in one go

```{r}
funimagery %>%
  group_by(intervention) %>%
  summarise(
    mean(weight_lost_end_trt),
    sd(weight_lost_end_trt)
  )

```



Give your new summary columns a name (this is good practice)

```{r}
funimagery %>%
  group_by(intervention) %>%
  summarise(
    mean_weight_lost_end_trt = mean(weight_lost_end_trt),
    sd_weight_lost_end_trt = sd(weight_lost_end_trt)
  )
```


# Boxplot

Boxplots show the interquartile range (IQR) as the height of the box.

The IQR is the range which includes 50% of the data points


```{r}
funimagery %>%
  ggplot(aes(intervention, weight_lost_end_trt)) +
  geom_boxplot() +
  scale_y_continuous(n.breaks = 10) + # this extra code just adds more marks on the y-axis
  labs(x = "Intervention", y="Weight lost (end treatment)")
```



# Scatter plot

```{r}
fuel %>%
  ggplot(aes(x=weight, y=mpg)) + # selects the columns to use
    geom_point()                 # adds the points to the plot
```







# Correlation (and Bayes Factor)

Load the `corrr` package (see above):

```{r}
# select weight at baseline and follow-up in the funimagery study
# and correlate them
funimagery %>%
  select(kg1, kg2, kg3) %>%
  correlate()
```


```{r}
# use `with()` to tell `correlationBF` which dataset to use
with(funimagery,  correlationBF(kg1, kg3))
```

Remember to ignore the part which reads `r=0.333`. The large number after that is the Bayes Factor in favour of the correlation.



# Loading CSV data


Load CSV data from a local file (in the same directory as your `.Rmd` file... remember to upload it to the server):


```{r, eval=F}
shootings <- read_csv("shootings.csv")
```



Load CSV data from a URL (over the web):

```{r}
shootings <- read_csv("https://raw.githubusercontent.com/benwhalley/psydata/main/data-raw/shootings.csv")
```




# Regression


```{r}
library(psydata)
```

Simple linear model (regression) with one predictor:

```{r}
lm(mpg ~ weight, data=fuel)
```

Calculate Bayes Factor for single predictor:

```{r}
library(BayesFactor)
```

```{r}
lmBF(mpg ~ weight, data=fuel)
```



Multiple regression (use `+` to add predictors:

```{r}
lm(mpg ~ weight + cyl, data=fuel)
```


Bayes Factor to test addition of a variable to an existing model:

```{r}
A <- lmBF(mpg ~ weight, data=fuel)
B <- lmBF(mpg ~ weight + cyl, data=fuel)

# gives evidence _for_ B against A
B / A
```




Calculate R^2^:

```{r}
library(broom)
lm(mpg ~ weight + cyl, data=fuel) %>%  glance()
```

Plot histogram of residuals from a regression model:

```{r}
lm(mpg ~ weight + cyl, data=fuel) %>% 
  augment() %>% 
  ggplot(aes(.resid)) + geom_histogram()
```


Fitted-vs-residual diagnostic plot:

```{r}
lm(mpg ~ weight + cyl, data=fuel) %>% 
  augment() %>% 
  ggplot(aes(.fitted, .resid)) + 
  geom_point() + 
  geom_smooth(size=.25) + 
  geom_smooth(method=lm, se=F, size=.25, color="black", linetype="dashed") + 
  theme_minimal()

```



